### 2011-11-07 doc/pldi2012/exec.tex
text cleanup, compaction

In section \ref{sec:type} we showed that the type system
only enables privileges to be passed with task calls.  Based on
this property of privileges we have claimed that our runtime
will be able to make local scheduling decisions.  In this
section we present the Legion execution model and show that
this property of scheduling preserves the sequential execution 
semantics of a program.  The correctness argument is
instructive in that the properties that ensure correctness are also
exploited in the implementation of the Legion high-level runtime
system to achieve high performance and highly distributed scheduling
of parallel tasks.

@@

There are four kinds of names: functions $f$, regions $r$, local
variables $x$, and partitions $p$.  A function takes only regions as
arguments and a function body consists of an ordered list of
statements.  The core language has no permissions or coherence annotations
on functions; for simplicity, in this section we simply use exclusive access in the case
that a function writes a region.  The statement $f(r_1,\ldots,r_n)$ calls a function, which
is executed for its effects on the region arguments.  The statement $x
= \deref r$ reads a value from region $r$ and stores it in local
variable $x$, and the statement $\deref r = x$ stores the value of $x$
in region $r$.

The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$ creates two
{\em subregions} $r_1$ and $r_2$ partitioning the parent region $r$.
All three region names ($r_1$,$r_2$,and $r$) can be used after the
partition.  Regions $r_1$ and $r_2$ are disjoint, meaning that writes
to $r_1$ have no effect on the contents of $r_2$ and vice versa, but
neither $r_1$ nor $r_2$ is disjoint from $r$.  The partition itself
also has a name $p$.  We define a relation $\rleq$ which is the
reflexive, transitive closure of
\[ r_1 \rleq p \ \ \ r_2 \rleq p \ \ \ p \rleq r \]
The $\rleq$ relation defines a {\em region forest} with regions at the
roots, and alternating levels where partitions are always children of
regions and regions are always children of partitions.  If we ignore
partitions, $\rleq$ is just the subregion relationship: $r \rleq r'$
means that $r$ is a subregion (either immmediately or transitively) of
$r'$.  Including partitions also gives information about disjointness.
Let $\lca{r}{r'}$ be the meet of $r$ and $r'$ (the least-common
ancestor of $r$ and $r'$ in the region forest).  If $r \neq r'$ and
$\lca{r}{r'} = p$ then $r$ and $r'$ are disjoint, as $r$ and $r'$ must
be in distinct subregions of the partition $p$ and partitions
guarantee disjoint subregions.  If $\lca{r}{r'} = r''$ then either $r
\rleq r' = r''$ , $r \rleq r = r''$, or $r \leq p$, $r' \leq p'$, and
$p$ and $p'$ are two distinct partitions of $r''$.  In any of these
cases $r$ and $r'$ cannot be proven disjoint using $\rleq$.

>>>>

In section \ref{sec:type} we showed that the Legion type system
only enables privileges to be passed with task calls. 
 In this
section we present the Legion execution model and show that
the sequential execution 
semantics of a program can be preserved while making only local scheduling
decisions.
  The correctness argument is
instructive in that the properties that ensure correctness are also
exploited in the implementation of the Legion high-level runtime
system to achieve high performance and highly distributed scheduling
of parallel tasks.

@@

There are four kinds of names: functions $f$, regions $r$, local
variables $x$, and partitions $p$.  A function takes only regions as
arguments and a function body consists of an ordered list of
statements.  The core language has no permissions or coherence annotations
on functions; in this section we simply use exclusive access in all cases.
 The statement $f(r_1,\ldots,r_n)$ calls a function, which
is executed for its effects on the region arguments.  The statement $x
= \deref r$ reads a value from region $r$ and stores it in local
variable $x$, and the statement $\deref r = x$ stores the value of $x$
in region $r$.

The statement $(r_1, r_2)_{p} = \mbox{\tt partition}(r)$ creates two
{\em subregions} $r_1$ and $r_2$ partitioning the parent region $r$.
All three region names ($r_1$,$r_2$,and $r$) can be used after the
partition.  Regions $r_1$ and $r_2$ are disjoint, meaning that writes
to $r_1$ have no effect on the contents of $r_2$ and vice versa, but
neither $r_1$ nor $r_2$ is disjoint from $r$.  The partition itself
also has a name $p$.  We define a relation $\rleq$ which is the
reflexive, transitive closure of
\[ r_1 \rleq p \ \ \ r_2 \rleq p \ \ \ p \rleq r \]
The $\rleq$ relation defines a {\em region forest} with regions at the
roots, and alternating levels where partitions are always children of
regions and regions are always children of partitions.  If we ignore
partitions, $\rleq$ is just the subregion relationship: $r \rleq r'$
means that $r$ is a subregion (either immmediately or transitively) of
$r'$.  Including partitions also gives information about disjointness.
Let $\lca{r}{r'}$ be the meet of $r$ and $r'$ (the least-common
ancestor of $r$ and $r'$ in the region forest).  If 
$\lca{r}{r'} = p$ then $r$ and $r'$ are disjoint, as $r$ and $r'$ must
be in distinct subregions of the partition $p$ and partitions
guarantee disjoint subregions.  If $r$ and $r'$ meet at a region (i.e. $\lca{r}{r'} = r''$) then either one of $r$ or $r'$ is a subregion of the other, or they
lie in two distinct partitions of $r''$ and disjointness cannot be guaranteed.

### 2011-11-07 doc/pldi2012/type_system.tex
text cleanup, compaction

The key to both performance and correctness in Legion is the
accuracy of the region usage declarations.  The runtime places data and
schedules tasks based on these declarations and has no way to recover if a 
task attempts to access data outside the declared regions or
writes a region the runtime expected would only be read.
The purpose of the Legion type system is to statically check that region usage declarations are adhered to.
In this section we summarize the main features of the type system; due to space limitations we omit more than
we include.
Type judgments have the form:
\begin{center}
$\Gamma, \Phi_i, C_i \vdash e : T, \Phi_o, C_o$
\end{center}
In addition to tracking the types of variables $\Gamma$, the
type environment includes the current privileges $\Phi_i$ 
and region constraints $C_i$.  A judgment concludes $e$ has some type $T$,
but also tells us how the evaluation of $e$ changes the privileges $\Phi_o$ and constraints $C_o$.

@@

This rule says that to read from a pointer, we must 
provide a region ($r_2$ in this case) for which we possess a read privilege that 
is known to either be, or contain, the region to which the pointer is
constrained.  The rule for region writes is similar.

Region relationships are first-class values and so can be stored in data structures
and passed around the program.  However, the Legion type
system cannot afford to lose information about regions that are stored
in the heap.  To track the regions
in region relationships accurately we use existential quantification
over regions.  The quantifiers capture both types and constraints, but 
not privileges.  The type of a region relation has the form:
\begin{center}
$RR = \exists r_1, \ldots, r_n.\left(T, \emptyset, C\right)$
\end{center}
Pack and unpack operations are used in the type system
to add and remove the existential quantifiers:
\begin{center}
{\small
\begin{math}
\begin{array}{c}
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : RR, \Phi_3, C_3  \\
r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}\left(\Gamma, \Phi_3, C_3\right)
\end{array}
}{
\begin{array}{l@{}l}
\Gamma, \Phi_2, C_2 \vdash unpackrr{ }e :  & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
& C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1
\end{array}
}
\\
\\
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, C_3 \\
C_3 \models [r'_1/r_1,\ldots,r'_n/r_n]C_1
\end{array}
}{
\begin{array}{l}
\Gamma, \Phi_2, C_2 \vdash packrr{ }RR\mbox{ }e : RR, \Phi_3, C_3 \\
\end{array}
}
\end{array}
\end{math}
}
\end{center}
Any expression whose type is a region relationship can be unpacked.
Doing so introduces a fresh region variable into the current type
environment for each region that was bound in the quantification.  The
fresh region variables are substituted into the region relationship's
type and constraints.  Packing an expression into a region
relationship does the reverse.  If the type of an expression can be
unified with the region relationship's type and the corresponding
constraints of the region relationship can be shown to hold, the
expression can be packed into the region relationship and used
interchangeably with any other instance of that region relationship.

Although the type system uses explicit packing and unpacking operations, 
note that these are implicit in Legion application code, such as the
circuit simulation in Listing~\ref{lst:code_ex}.  Unpacking is
inserted whenever a variable comes into scope, and packing is used
whenever a variable is written to the heap or passed as a task argument.

@@

As mentioned above, region privileges are not stored in packed types;
instead, privileges are passed through the task call tree.  This is a key property of the type
system, and is what allows the runtime to 
track which regions are in use by which tasks and to make distributed scheduling decisions.
Each subtask gets the privileges it requires from
the task that calls it.  For example, in Listing~\ref{lst:code_ex},
the task {\tt calc\_new\_currents} takes regions {\tt piece.rw\_ovt} with {\tt RW} privileges
and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges
as arguments.  Because the calling task has {\tt RW} privileges on the parent regions of these
region arguments, the type checker confirms that the subtask can execute with the needed privileges.
The coherence property for a task argument is always identical to the one of the parent task.

Because privileges are forgotten when a region relationship is packed, one may wonder how
a task can ever actually use a region $r$ in the region relationship when it is unpacked.
The answer is that if $r \rleq r'$ and the task already holds privileges for $r'$, then it is able
to use $r'$ with those same privileges.

>>>>

The key to both performance and correctness in Legion is the
accuracy of the region usage declarations.  The Legion type system has been
designed to statically check that region usage declarations are adhered to.
In this section we summarize the main features of the type system; due to space limitations we omit more than
we include.
Type judgments have the form:
\begin{center}
$\Gamma, \Phi_i, C_i \vdash e : T, \Phi_o, C_o$
\end{center}
In addition to tracking the types of variables $\Gamma$, the
type environment includes the current privileges $\Phi_i$ 
and region constraints $C_i$.  A judgment concludes $e$ has some type $T$,
but also tells us how the evaluation of $e$ changes the privileges $\Phi_o$ and constraints $C_o$.

@@

In order to safely read from a pointer into $r_1$, we must 
provide a region $r_2$ which is known to contain $r_1$ and for which we possess
a read privilege.  Rules for region writes and reductions are similar.

Region relationships are first-class values and can be stored in data structures
and passed around the program.  However, the Legion type
system cannot afford to lose information about regions that are stored
in the heap.  To track the regions
in region relationships accurately we use existential quantification
over regions.  The quantifiers capture both types and constraints, but 
not privileges.  The type of a region relation has the form:
\begin{center}
$RR = \exists r_1, \ldots, r_n.\left(T, \emptyset, C\right)$
\end{center}
Pack and unpack operations are used in the type system
to add and remove the existential quantifiers:
\begin{center}
{\small
\begin{math}
\begin{array}{c}
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : RR, \Phi_3, C_3  \\
r'_1, \ldots, r'_n \not\in \mathit{RegionsOf}\left(\Gamma, \Phi_3, C_3\right)
\end{array}
}{
\begin{array}{l@{}l}
\Gamma, \Phi_2, C_2 \vdash unpackrr{~}e :~  & [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, \\
& C_3 \wedge [r'_1/r_1,\ldots,r'_n/r_n]C_1
\end{array}
}
\\
\\
\infrule{
\begin{array}{l}
RR = \exists r_1, \ldots, r_n.\left(T_1, \emptyset, C_1\right) \\
\Gamma, \Phi_2, C_2 \vdash e : [r'_1/r_1,\ldots,r'_n/r_n]T_1, \Phi_3, C_3 \\
C_3 \models [r'_1/r_1,\ldots,r'_n/r_n]C_1
\end{array}
}{
\begin{array}{l}
\Gamma, \Phi_2, C_2 \vdash packrr{ }RR\mbox{ }e : RR, \Phi_3, C_3 \\
\end{array}
}
\end{array}
\end{math}
}
\end{center}
Any expression whose type is a region relationship can be unpacked.
Doing so introduces a fresh region variable into the current type
environment for each region that was bound in the quantification.  The
fresh region variables are substituted into the region relationship's
type and constraints.  Packing an expression into a region
relationship does the reverse.  If the type of an expression can be
unified with the region relationship's type and the corresponding
constraints of the region relationship can be shown to hold, the
expression can be packed into the region relationship and used
interchangeably with any other instance of that region relationship.

Although the type system uses explicit packing and unpacking operations, 
these are implicit in Legion application code, such as the
circuit simulation in Listing~\ref{lst:code_ex}.  Unpacking is
inserted whenever a variable comes into scope, and packing is used
whenever a variable is written to the heap or passed as a task argument.

@@

As mentioned above, region privileges are not stored in packed types;
instead, privileges are passed through the task call tree.  This is a key
property of the type system, and is what allows the runtime to 
track which regions are in use by which tasks and to make distributed
scheduling decisions.
Each subtask must obtain its privileges from
the task that calls it.  For example, in Listing~\ref{lst:code_ex},
the task {\tt calc\_new\_currents} requires regions {\tt piece.rw\_pvt} with {\tt RW} privileges
and {\tt piece.rn\_pvt} and {\tt piece.rn\_ghost} with {\tt RO} (read-only) privileges.  Because the calling task has {\tt RW} privileges on the parent regions of these
region arguments, the type checker is able to confirm that the subtask can execute with the needed privileges.

%The coherence property for a task argument is always identical to the one of the parent task.

Due the the inability for a packed region relationship to hold privileges, any
useful region relationship will leave at least one region free (i.e. unquantified) and include constraints that relate the quantified region(s) to the free one(s).  As long as a task holds a privilege for the region that is kept free, the
region relationship can be unpacked and used.
